A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-09-09 submission/manuscript.rmd
modify results section to demonstrate power of OMOP

\captionsetup{width=\textwidth}
```{r table, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
table <- readr::read_csv(
  here::here("data/processed/table1.csv"),
  na = character()
) %>%
  select(-database)

@@

\captionsetup{width=\textwidth}
```{r tableAucs, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  population = c(
    "Matched set",
    "Treatment cohort",
    "Comparator cohort",
    "Entire population"
  ),
  ccae = c(
    75.78, 70.65, 79.1, 74.06
  ),
  mdcd = c(
    78.12, 78.29, 80.26, 79.46
  ),
  mdcr = c(
    71.09, 69.03, 68.53, 69.07
  )
)

@@

## Step 5: Result presentation and evaluation

We present the results of the analysis in the CCAE database with
stratification based on risk predictions of acute MI. Results of analyses in the
other databases and with other risk stratifications are included in the
supplementary material.

For each outcome and in each risk stratum there were adequate numbers of
patients (Table \@ref(tab:table)). The discriminative ability of the prediction
models was moderate in the matched development subset (c-index 0.76 for acute MI
; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
hospitalization with heart failure was 0.79 and for stroke it was 0.75).

Relative treatment effects of ACE-inhibitors vs beta blockers increased in favor
of ACE-inhibitors (hazard ratios decreased) with increasing acute MI risk,
resulting in more pronounced absolute risk differences (ARD) with increasing
acute MI risk (Figure \@ref(fig:figure2)).

```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI. The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk. These are derived using the KaplanMeier estimates at 730 days after inclusion. The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI. These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk. The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers. These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion. Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk. Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
```

In general, with increasing acute MI risk we observed an increasing pattern in
terms of absolute benefit for the main outcomes, while the absolute harms did
not increase for the safety outcomes. More specifically, patients in the low
risk quarter did not receive absolute benefit (-0.03%) while absolute risk
was 0.54% lower (95% confidence interval 0.36%—0.71%) for patients in the
highest risk quarter. In contrast, the absolute and relative effects of
ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are slightly
decreasing with increasing acute MI risk (Figure \@ref(fig:figure3) and
\@ref(fig:figure4)). Similar results were observed in the other two databases
(see supplementary material).

```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
```

```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
```

These results suggest that treatment with ACE-inhibitors, compared to treatment
with beta blockers, may be focused on the higher risk patients, in whom the
benefits outweigh the harms, while beta blockers may be a viable option in lower
risk patients, in whom the benefit-harm tradeoff is more favorable. This is in
accordance with earlier findings that beta blockers should also be considered as
first-line treatment for younger hypertensive patients [@Cruickshank2007;
@Khan2006]. More thorough evaluation of these results is required in future
research, however.

>>>>

\captionsetup{width=\textwidth}
```{r table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
table <- readr::read_csv(
  here::here("data/processed/table1.csv"),
  na = character()
) %>%
  select(-database)

@@

\captionsetup{width=\textwidth}
```{r tableAucs, cache=TRUE, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  population = c(
    "Matched set",
    "Treatment cohort",
    "Comparator cohort",
    "Entire population"
  ),
  ccae = c(
    75.78, 70.65, 79.1, 74.06
  ),
  mdcd = c(
    78.12, 78.29, 80.26, 79.46
  ),
  mdcr = c(
    71.09, 69.03, 68.53, 69.07
  )
)

@@

In general, there was sufficient overlap of propensity score distribution in all
risk strata (Figure \@ref(fig:psDensity)). If empirical equipoise was not
achieved, the validity of the comparative effectiveness estimates would be
questionable.

```{r psDensity, cache = TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk. The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
```

Propensity score adjustment achieved balance for most of the considered
covariates, measured using standardized mean differences before and after
adjustment (Figure \@ref(fig:covariateBalance)). However, in lower risk strata
imbalances persisted for a substantial subset of the covariates, all related to
pregnancy findings. This was anticipated (use of ACE inhibitors is specifically
contraindicated during pregnancy) and was already pointed out in
[@Suchard2019].

```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient charactersistic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores. Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
```

For hospitalization with acute MI there was an increasing trend in favor ACE
inhibitors compared to beta blockers on the relative scale (hazard ratios
decreased) with increasing acute MI risk. More specifically, hazard ratios
decreased from 0.98 (0.77 to 1.26; 95% CI), 1.30 (0.51 to 3.22; 95% CI) and 1.03
(0.82 to 1.29; 95% CI) to 0.76 (0.71 to 0.82; 95% CI), 0.94 (0.82 to 1.07; 95%
CI) and 1.03 (0.93 to 1.15; 95% CI) in CCAE, MDCD and MDCR respectively (Figure
\@ref(fig:combinedRelative)). In terms of hospitalization with heart failure
relative treatment effect estimates favored ACE inhibitors across all risk
strata in all databases. We found no differences between the two treatments in
their effect on stroke on the relative scale. In terms of the safety outcomes we
found an increased risk of cough and angioedema on the relative scale across all
risk strata. In the case of cough, this effect decreased with increasing risk of
acute MI---from 1.37 (1.33 to 1.41; 95% CI), 1.35 (1.24 to 1.48; 95% CI)
and 1.37 (1.29 to 1.45; 95% CI) in the lowest acute MI risk quarter to 1.26
(1.22 to 1.29; 95% CI), 1.07 (1.00 to 1.14) and 1.10 (1.04 to 1.17; 95% CI) in
the highest acute MI risk quarter in CCAE, MDCD and MDCR respectively.

```{r combinedRelative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI. Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative.tiff")))
```

We observed an increasing trend of treatment effect on the absolute scale with
increasing acute MI risk in favor of ACE inhibitors in terms of acute MI in all
databases except for MDCR---from -0.03% (-0.08% to 0.01%; 95% CI), -0.05%
(-0.18% to 0.08%; 95% CI) and -0.02% (-0.24% to 0.19%; 95% CI) in the lowest
acute MI risk quarter to 0.54% (0.36% to 0.71%; 95% CI), 0.29% (-0.39%
to 0.97%; 95% CI) and -0.39% (-0.96% to 0.18%; 95% CI) in the highest acute MI
risk quarter in CCAE, MDCD and MDCR, respectively (Figure \@ref(fig:figure12)).
We found no difference on the absolute scale for stroke across risk strata.
Absolute risk differences did not favor ACE inhibitors compared to beta blockers
in terms of cough, even though this effect again diminished with increasing
acute MI risk---from -4.14% (-4.62% to -3.66%; 95% CI), -6.45% (-9.12% to
-3.78%; 95% CI) and -4.81% (-5.76% to -3.85%; 95% CI) in the lowest acute MI
risk quarter to -2.57% (-2.99% to -2.15%; 95% CI), -1.11% (-2.93% to 0.70%; 95%
CI) and -1.69% (-2.83% to -0.55%; 95% CI) in the highest acute MI risk quarter
in CCAE, MDCD and MDCR, respectively. In terms of angioedema absolute risk
differences were very small due to the rarity of the outcome.

```{r figure12, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the absolute scale within strata of predicted risk of acute MI. Estimates of absolute treatment effect are derived as the difference in Kaplan-Meier estimates at 730 after inclusion. Values above 0 favor ACE inhibitors, while values below 0 favor beta blockers.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/CombinedAbsolute.tiff")))
```

These results suggest that treatment with ACE-inhibitors, compared to treatment
with beta blockers, may be focused on the higher risk patients, in whom the
benefits outweigh the harms, while beta blockers may be a viable option in lower
risk patients, in whom the benefit-harm tradeoff is more favorable. This is in
accordance with earlier findings that beta blockers should also be considered as
first-line treatment for younger hypertensive patients [@Cruickshank2007;
@Khan2006]. This analysis, however, was carried out as demonstration of the
framework and more rigorous analyses are required to make any suggestions for
clinical practice.
